%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE acknowledgements.tex}%

\begin{ntacknowledgements}

In November 2022, I embarked on a journey of over 10,000 kilometers from Santiago de Chile to Lisbon, driven by a desire to explore the immense potential of Geographic Information Systems. Over the course of these two years, I have had the privilege of living in a country that embraced me with warmth and hospitality, allowing me to grow academically, personally, and professionally. Along the way, I also met incredible people who made this journey unforgettable.

First and foremost, I would like to express my deepest gratitude to Professor Marco Painho, the coordinator of the Master's program, for his guidance and leadership throughout this academic journey. I would also like to thank Professor Pedro Cabral, my advisor, for his support in supervising this thesis.

I am deeply grateful to the entire team of NOVA IMS professors for their dedication and for providing a solid foundation of knowledge that made this program a rewarding and enriching experience.

To Gabriel, thank you for your endless patience, encouragement, and support throughout this journey. Your presence has been my anchor, giving me the strength to persevere during challenging times.

To my parents, Mauricio and Ana María, thank you for your unwavering support, even from thousands of kilometers away in Chile. Your love and belief in me have been a constant source of motivation, and I couldn't have done this without you.

Finally, I am deeply thankful to this land that welcomed me with open arms and gave me the opportunity to thrive and explore my potential. This thesis is the culmination of not only my academic efforts but also the collective contributions, love, and support of the people around me. To all of you, I am profoundly grateful. Thank you for being part of this journey.

\end{ntacknowledgements}